Analysis of Tick-Bourne Disease Cases Reported to DPH Between 2019-2024

### 2025-05-03 products/manuscript/Manuscript.qmd
Final Project Push

---
title: "Analysis of Tick-Bourne Disease Cases Reported to DPH Between 2019-2024"
author: ""
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/american-journal-of-epidemiology.csl
editor: 
  markdown: 
    wrap: 72
---

@@

This project explores patterns in tick-borne disease cases reported to
the Georgia Department of Public Health (DPH) between 2019 and 2024. The
analysis involved extensive data cleaning to retain only confirmed and
positive cases, with additional processing to resolve inconsistencies in
timelines and reporting. Exploratory data analysis (EDA) was used to
examine disease distribution, demographic patterns, geographic hotspots,
and data completeness over time. Building on these insights, predictive
models were developed to assess whether patient and case-level
factors—such as age, disease type, and county—could help identify cases
at risk of delayed follow-up. This work theorectically supports improved
understanding of disease surveillance data and highlights areas for
strengthening tick-borne disease reporting and case management systems
in Georgia.

@@

### 2.1 Background

In the past two decades, the study of tick-borne disease has gained
increasing attention due to rising incidence rates and expanding
geographic ranges of disease-carrying ticks. While early medical and
veterinary research in the 20th century identified key tick-borne
illnesses such as Rocky Mountain spotted fever and Lyme disease, recent
decades have seen an acceleration in the detection of new pathogens,
greater public health concern, and expanded surveillance efforts across
the globe. Researchers have documented the ecology and behavior of
ticks, explored the role of climate and land use in tick habitat
expansion, and analyzed patterns of disease incidence in both human and
animal populations. Despite the growing body of research, many knowledge
gaps remain, particularly in understanding co-infections, long-term
health effects, and effective prevention strategies. Tick-borne diseases
(TBDs) are primarily transmitted through the bite of infected
hard-bodied ticks (family Ixodidae), and include bacterial, viral, and
protozoan pathogens. Today, TBDs are recognized as a major public health
concern in both rural and suburban settings, with transmission risk
influenced by ecological, climatic, and sociocultural factors.

### 2.1.1 General Background Information

Tick-borne illnesses are diseases that humans can contract from the bite
of an infected tick. Ticks are small, blood-sucking arachnids that are
often found in wooded, grassy, or suburban areas. When a tick feeds on
an infected animal, it can acquire pathogens—such as bacteria, viruses,
or parasites—that can then be transmitted to humans. Examples of
tick-borne diseases include, but are not limited to, Lyme disease,
Anaplasmosis, and Babesiosis. These diseases are of increasing public
health concern, especially as the geographic range and seasonality of
ticks expand due to environmental and ecological changes.

### 2.1.2 Transmission and Reservoirs

Ticks act as both vectors and reservoirs for a variety of pathogens,
transmitting disease to humans and animals during blood meals. Most
tick-borne diseases are zoonotic, with small mammals, deer, and birds
acting as key reservoirs in natural cycles. Transmission to humans
typically occurs when a nymph or adult tick feeds for an extended
period, allowing the pathogen to transfer from the tick’s salivary
glands into the host’s bloodstream. Common tick-borne pathogens include
Borrelia burgdorferi (the agent of Lyme disease), Rickettsia rickettsii
(Rocky Mountain spotted fever), Anaplasma phagocytophilum, Ehrlichia
chaffeensis, and viruses such as Powassan and Heartland virus. Ticks
themselves are influenced by multiple environmental factors, including
temperature, humidity, vegetation, and host availability. Human exposure
risk is increased in areas with high host density, edge habitats between
forest and residential zones, and during seasons with peak tick
activity. Additionally, land use changes—such as suburban development,
deforestation, and agricultural expansion—can modify tick habitats and
affect pathogen dynamics.

### 2.1.3 Ecological and Health Impacts

The ecological consequences of tick-borne disease extend beyond human
health. Tick populations and their associated pathogens can alter
wildlife health and community composition. Infected animals may suffer
sublethal effects that reduce fitness, while high burdens of tick
parasitism can influence population dynamics, particularly in rodents
and ungulates. From a public health perspective, TBDs can cause a range
of acute and chronic symptoms in humans, from fever and fatigue to
long-term neurological and musculoskeletal complications.
Co-infections—when an individual is infected with more than one
pathogen—are increasingly recognized and may complicate diagnosis and
treatment. The public health burden is also shaped by disparities in
access to health care, awareness of TBDs, and availability of diagnostic
testing, which may result in underreporting or misdiagnosis. The costs
associated with tick-borne disease—including medical care, productivity
loss, and prevention efforts—represent a growing concern for health
systems, particularly in endemic areas.

### 2.1.4 Relevance

Tick-borne diseases have been documented in all regions of the United
States, with the southeastern, northeastern, and upper midwestern states
experiencing particularly high incidence rates. Climate models suggest
that warming temperatures and shifting precipitation patterns may expand
the range of ticks and lengthen their active season, potentially
increasing the geographic footprint and seasonality of tick-borne
disease transmission. In the southeastern U.S., for example, Amblyomma
americanum (the lone star tick) has expanded in abundance and geographic
range over the past decades, contributing to increased cases of
ehrlichiosis and the emergence of Alpha-Gal Syndrome, a red meat allergy
triggered by tick bites. Understanding the ecology, distribution, and
drivers of tick-borne disease is therefore crucial to predicting future
trends, identifying at-risk populations, and developing effective
prevention and response strategies.

# 3. The Data

# 3.1 Data Acquisition & Description

The data used in this analysis was obtained through my internship with
the Georgia Department of Health, in collaboration with the Centers for
Disease Control and Prevention (CDC). Specifically, it is derived from
the Nationally Notifiable Diseases Surveillance System (NNDSS) and
focuses on tick-borne illnesses. This data set provides key insights
into the trends, distribution, and characteristics of reported cases,
supporting public health efforts in disease monitoring and
prevention.2.1.5 Description of Data and Data Source

@@

## 3.1.2 Questions and Hypotheses to Be Addressed

@@

# 4. Methods

## 4.1 Workflow Summarized

@@

## 4.2 Uncleaned Data

@@

## 4.3 Initial Data Cleaning

@@

#### 4.3.1 Looked for Missing Data & General Trends

@@

## 4.4 Exploratory/Descriptive Analysis

#### 4.4.1 Identified Disease Burden Trends & Hotspot Counties

@@

#### 4.4.2 Checked Age Distribution

@@

#### 4.4.3 Evaluated Data Completeness Over Time

@@

#### 4.4.4 Analyzed Timeline Between Key Case Events

@@

## 4.5 EDA Revelant Takeaways

#### 4.5.1 Most Common Tick-Borne Diseases

The dataset is filtered to identify the five most frequently reported
tick-borne diseases. A bar chart visualizes their case counts, offering
an overview of significant public health concerns.

@@

#### 4.5.2 Geographic Distribution

@@

#### 4.5.3 Data Completeness Over Time

@@

@@

#### 4.5.4 Timeline Analysis of Key Case Events

@@

## 4.6 Statistical Analysis

@@

# 5. Results

## 5.1 Basic statistical analysis

@@

+-----------+-----------+-----------+-----------+-----------+
| term      | estimate  | s         | statistic | p.value   |
|           |           | td.error  |           |           |
| \<chr\>   | \<dbl\>   |           | \<dbl\>   | \<dbl\>   |
|           |           | \<dbl\>   |           |           |
+:==========+==========:+==========:+==========:+==========:+
| ( I       | -2 3 2    | 1 57      | 1         | 0.1729390 |
| ntercept) | 8.5151515 | 3.052665  | .48025251 |           |
+-----------+-----------+-----------+-----------+-----------+
| YEAR      | 1.1515152 | 0.777581  | 1         | 0.1727716 |
|           |           |           | .48089418 |           |
+-----------+-----------+-----------+-----------+-----------+
| B         | 0.4242424 | 5.484565  | 0         | 0.9400358 |
| ABESIOSIS |           |           | .07735207 |           |
+-----------+-----------+-----------+-----------+-----------+
| DENGUE    | 4.8484848 | 6.364775  | 0         | 0.4656907 |
|           |           |           | .76176850 |           |
+-----------+-----------+-----------+-----------+-----------+
| EHRLICHIA | 5.0000000 | 5.157889  | 0         | 0.3576794 |
| C H       |           |           | .96938889 |           |
| AFFEENSIS |           |           |           |           |
+-----------+-----------+-----------+-----------+-----------+
| EHRLICHIA | -         | 6.364775  | 0         | 0.8604390 |
| EWINGII   | 1.1515152 |           | .18092002 |           |
+-----------+-----------+-----------+-----------+-----------+
| LYME      | 8.5606061 | 4.963745  | 1         | 0.1186814 |
| DISEASE   |           |           | .72462659 |           |
+-----------+-----------+-----------+-----------+-----------+
| ROCKY     | 0.5757576 | 5.484565  | 0         | 0.9186958 |
| MOUNTAIN  |           |           | .10497781 |           |
| SPOTTED   |           |           |           |           |
| FEVER     |           |           |           |           |
+-----------+-----------+-----------+-----------+-----------+
| WEST NILE | 8.2272727 | 5.593721  | 1         | 0.1754208 |
| (WNV)     |           |           | .47080497 |           |
| INFECTION |           |           |           |           |
+-----------+-----------+-----------+-----------+-----------+

@@

@@

+--------+--------+--------+--------+--------+
| term   | e s    | s t d  | s t a  | p      |
|        | timate | .error | tistic | .value |
| \      |        |        |        |        |
| \      | \      | \      | \      | \      |
| <chr\> | \      | \      | \      | \      |
|        | <dbl\> | <dbl\> | <dbl\> | <dbl\> |
+:=======+=======:+=======:+=======:+=======:+
| ( I n  | 9 9 3  | 5 7 2  | 1 .    | 0 . 1  |
| t e    | 5 .    | 9 .    | 733920 | 434648 |
| rcept) | 267857 | 946482 |        |        |
+--------+--------+--------+--------+--------+
| YEAR   | -   4  | 2 .    | -   1  | 0 . 1  |
|        |     .  | 835202 |     .  | 438108 |
|        |        |        |        |        |
|        | 910714 |        | 732051 |        |
+--------+--------+--------+--------+--------+

@@

## 5.2 Further Statistical Analysis

@@

## 5.3 Adding in Poverty Rate Data by County

@@

+--------------+--------------+--------------+--------------+
|              |              |              |              |
+--------------+--------------+--------------+--------------+
|              |              |              |              |
+--------------+--------------+--------------+--------------+

@@

# 6. Discussion

## 6.1 Summary and Interpretation

@@

The ROC curve showed consistent separation between sensitivity and
1-specificity, further supporting model validity. These results suggest
that case characteristics such as age, sex, county, and disease type
contain meaningful predictive information regarding follow-up delays in
surveillance data.Such insights can inform improvements in case
monitoring, triaging, and outreach in counties or patient groups with
higher risk for delays.

County-level poverty rate was evaluated as a predictor of the proportion
of cases with diagnostic delays exceeding 30 days. A linear regression
model indicated no statistically significant relationship between
poverty rate and delay rate (p = 0.87). The estimated effect size was
minimal (β = 0.0049), and the 95% confidence interval included zero.

@@

## 6.2 Strengths and Limitations

@@

# 7. Conclusions

>>>>

---
title: "Analysis of Tick-Bourne Disease Cases Reported to DPH Between 2019-2024"
author: ""
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/references.bib
csl: ../../assets/vancouver.csl
editor: 
  markdown: 
    wrap: 72
---

@@

This project explores patterns in tick-borne disease cases reported to
the Georgia Department of Public Health (DPH) between 2019 and 2024. The
analysis involved extensive data cleaning to retain only confirmed and
positive cases, with a\@perumalsamy2024dditional processing to resolve
inconsistencies in timelines and reporting. Exploratory data analysis
(EDA) was used to examine disease distribution, demographic patterns,
geographic hotspots, and data completeness over time. Building on these
insights, predictive models were developed to assess whether patient and
case-level factors—such as age, disease type, and county—could help
identify cases at risk of delayed follow-up. This work theorectically
supports improved understanding of disease surveillance data and
highlights areas for strengthening tick-borne disease reporting and case
management systems in Georgia.

@@

### Background

In the past two decades, the study of tick-borne disease has gained
increasing attention due to rising incidence rates and expanding
geographic ranges of disease-carrying ticks. While early medical and
veterinary research in the 20th century identified key tick-borne
illnesses such as Rocky Mountain spotted fever and Lyme disease, recent
decades have seen an acceleration in the detection of new pathogens,
greater public health concern, and expanded surveillance efforts across
the globe [@mcpherson2017]. Researchers have documented the ecology and
behavior of ticks, explored the role of climate and land use in tick
habitat expansion, and analyzed patterns of disease incidence in both
human and animal populations. Despite the growing body of research, many
knowledge gaps remain, particularly in understanding co-infections,
long-term health effects, and effective prevention strategies
[@usepa2016]. Tick-borne diseases (TBDs) are primarily transmitted
through the bite of infected hard-bodied ticks (family Ixodidae), and
include bacterial, viral, and protozoan pathogens. Today, TBDs are
recognized as a major public health concern in both rural and suburban
settings, with transmission risk influenced by ecological, climatic, and
sociocultural factors [@perumalsamy2024].

### General Background Information

Tick-borne illnesses are diseases that humans can contract from the bite
of an infected tick. Ticks are small, blood-sucking arachnids that are
often found in wooded, grassy, or suburban areas. When a tick feeds on
an infected animal, it can acquire pathogens—such as bacteria, viruses,
or parasites—that can then be transmitted to humans. Examples of
tick-borne diseases include, but are not limited to, Lyme disease,
Anaplasmosis, and Babesiosis. These diseases are of increasing public
health concern, especially as the geographic range and seasonality of
ticks expand due to environmental and ecological changes @cdc2024a.

### Transmission and Reservoirs

Ticks act as both vectors and reservoirs for a variety of pathogens,
transmitting disease to humans and animals during blood meals. Most
tick-borne diseases are zoonotic, with small mammals, deer, and birds
acting as key reservoirs in natural cycles. Transmission to humans
typically occurs when a nymph or adult tick feeds for an extended
period, allowing the pathogen to transfer from the tick’s salivary
glands into the host’s bloodstream. Common tick-borne pathogens include
Borrelia burgdorferi (the agent of Lyme disease), Rickettsia rickettsii
(Rocky Mountain spotted fever), Anaplasma phagocytophilum, Ehrlichia
chaffeensis, and viruses such as Powassan and Heartland virus. Ticks
themselves are influenced by multiple environmental factors, including
temperature, humidity, vegetation, and host availability[@berger2014;
@bouchard2019]. Human exposure risk is increased in areas with high host
density, edge habitats between forest and residential zones, and during
seasons with peak tick activity [@lerman2018; @aronson2017].
Additionally, land use changes—such as suburban development,
deforestation, and agricultural expansion—can modify tick habitats and
affect pathogen dynamics @allan2003.

### Ecological and Health Impacts

The ecological consequences of tick-borne disease extend beyond human
health. Tick populations and their associated pathogens can alter
wildlife health and community composition. Infected animals may suffer
sublethal effects that reduce fitness, while high burdens of tick
parasitism can influence population dynamics, particularly in rodents
[@kasari2010; @clark]. From a public health perspective, TBDs can cause
a range of acute and chronic symptoms in humans, from fever and fatigue
to long-term neurological and musculoskeletal complications
@zychowski2024. Co-infections—when an individual is infected with more
than one pathogen—are increasingly recognized and may complicate
diagnosis and treatment. The public health burden is also shaped by
disparities in access to health care, awareness of TBDs, and
availability of diagnostic testing, which may result in underreporting
or misdiagnosis. The costs associated with tick-borne disease—including
medical care, productivity loss, and prevention efforts—represent a
growing concern for health systems, particularly in endemic areas
@berghoff2012.

### Relevance

Tick-borne diseases have been documented in all regions of the United
States, with the southeastern, northeastern, and upper midwestern states
experiencing particularly high incidence rates. Climate models suggest
that warming temperatures and shifting precipitation patterns may expand
the range of ticks and lengthen their active season, potentially
increasing the geographic footprint and seasonality of tick-borne
disease transmission [@gage2008; @dawe2016]. In the southeastern U.S.,
for example, Amblyomma americanum (the lone star tick) has expanded in
abundance and geographic range over the past decades, contributing to
increased cases of ehrlichiosis and the emergence of Alpha-Gal Syndrome,
a red meat allergy triggered by tick bites @cdc2022. Understanding the
ecology, distribution, and drivers of tick-borne disease is therefore
crucial to predicting future trends, identifying at-risk populations,
and developing effective prevention and response strategies.

# The Data

# Data Acquisition & Description

The data used in this analysis was obtained through my internship with
the Georgia Department of Health, in collaboration with the Centers for
Disease Control and Prevention (CDC). Specifically, it is derived from
the Nationally Notifiable Diseases Surveillance System (NNDSS) and
focuses on tick-borne illnesses. This data set provides key insights
into the trends, distribution, and characteristics of reported cases,
supporting public health efforts in disease monitoring and prevention.

@@

## Questions and Hypotheses to Be Addressed

@@

# Methods

## Workflow Summarized

@@

## Uncleaned Data

@@

## Initial Data Cleaning

@@

#### Looked for Missing Data & General Trends

@@

## Exploratory/Descriptive Analysis

#### Identified Disease Burden Trends & Hotspot Counties

@@

#### Checked Age Distribution

@@

#### Evaluated Data Completeness Over Time

@@

#### Analyzed Timeline Between Key Case Events

@@

## EDA Relevant Takeaways

#### Most Common Tick-Borne Diseases

The data set is filtered to identify the five most frequently reported
tick-borne diseases. A bar chart visualizes their case counts, offering
an overview of significant public health concerns.

@@

**Figure 1.** Distribution of tick-borne disease case counts in Georgia,
2019–2024.

#### Geographic Distribution

@@

**Figure 2-4.** The distribution of the most common disease types across
all counties– narrowed to the counties with more tick cases (hotspot
counties) then further narrowed focus on the top five tick-borne
diseases within these hotspot counties

#### Data Completeness Over Time

@@

**Figure 5.** Data completeness variation in the data set and entry
consistency over time

@@

#### Timeline Analysis of Key Case Events

@@

## Statistical Analysis

@@

# Results

## Basic statistical analysis

@@

+----------------+-------------+-----------+-----------+-----------+
| term\          | estimate\   | std.      | st        | p.value\  |
| \<chr\>        | \<dbl\>     | error\    | atistic\  | \< dbl\>  |
|                |             | \<dbl\>   | \<dbl\>   |           |
+:===============+============:+==========:+==========:+==========:+
| (Intercept)    | \-          | 157       | 1.        | 0         |
|                | 2           | 3.052665  |  48025251 | .1729390  |
|                | 328.5151515 |           |           |           |
+----------------+-------------+-----------+-----------+-----------+
| YEAR           | 1.1515152   | 0.777581  | 1.        | 0         |
|                |             |           |  48089418 | .1727716  |
+----------------+-------------+-----------+-----------+-----------+
| BABESIOSIS     | 0.4242424   | 5 .484565 | 0 .       | 0         |
|                |             |           | 07735207  | .9400358  |
+----------------+-------------+-----------+-----------+-----------+
| DENGUE         | 4.8484848   | 6\.       | 0 .       | 0         |
|                |             | 364775    | 76176850  | .4656907  |
+----------------+-------------+-----------+-----------+-----------+
| EHRLICHIA      | 5.0000000   | 5.157889  | 0 .       | 0         |
| CHAFFEENSIS    |             |           | 96938889  | .3576794  |
+----------------+-------------+-----------+-----------+-----------+
| EHRLICHIA      | \- 1 .      | 6.364775  | 0 .       | 0         |
| EWINGII        | 1515152     |           | 18092002  | .8604390  |
+----------------+-------------+-----------+-----------+-----------+
| LYME DISEASE   | 8.5606061   | 4.963745  | 1 .       | 0         |
|                |             |           | 72462659  | .1186814  |
+----------------+-------------+-----------+-----------+-----------+
| ROCKY MOUNTAIN | 0.5757576   | 5.484565  | 0\.       | 0         |
| SPOTTED FEVER  |             |           | 10497781  | .9186958  |
+----------------+-------------+-----------+-----------+-----------+
| WEST NILE      | 8.2272727   | 5.593721  | 1.        | 0         |
| (WNV)          |             |           |  47080497 | .1754208  |
| INFECTION      |             |           |           |           |
+----------------+-------------+-----------+-----------+-----------+

@@

**Table 1.** Output form a linear model predicting total case counts by
year and disease type. The intercept represents baseline case counts
while each diseases represents additional effect size.

@@

+---------+---------+---------+---------+---------+
| term    | e       | std .   | s t a t | p .     |
|         | stimate | error   | istic   | value   |
| \       |         |         |         |         |
| \       | \       | \       | \       | \       |
| \<chr\> | \       | \       | \       | \       |
|         | \<      | \<      | \<      | \<      |
|         | dbl\>   | dbl\>   | dbl\>   | dbl\>   |
+:========+========:+========:+========:+========:+
| ( I n   | 9935 .  | 5729 .  | 1 .     | 0 .     |
| t       | 267857  | 946482  | 733920  | 1434648 |
| ercept) |         |         |         |         |
+---------+---------+---------+---------+---------+
| YEAR    | 4 .     | 2 .     | 1 .     | 0 .     |
|         | 910714  | 835202  | 732051  | 1438108 |
+---------+---------+---------+---------+---------+

@@

## Further Statistical Analysis

@@

## Adding in Poverty Rate Data by County

@@

@@

# Discussion

## Summary and Interpretation

@@

The ROC curve showed consistent separation between sensitivity and
1-specificity, further supporting model validity. These results suggest
that case characteristics such as age, sex, county, and disease type
contain meaningful predictive information regarding follow-up delays in
surveillance data. These results align with previous findings that
individual-level demographics and geography can influence diagnosis of
vector-borne diseases, though the moderate predictive power suggests
additional factors— such as healthcare access and proximity
[@starke2023; @frontier]— remain unmeasured in this simplistic modeled
understanding of time to diagnosis. Such insights can inform
improvements in case monitoring, triaging, and outreach in counties or
patient groups with higher risk for delays.

County-level poverty rate was evaluated as a predictor of the proportion
of cases with diagnostic delays exceeding 30 days. A linear regression
model indicated no statistically significant relationshipbetween poverty
rate and delay rate (p = 0.87). The estimated effect size was minimal (β
= 0.0049), and the 95% confidence interval included zero.

@@

## Strengths and Limitations

@@

# Conclusions
